---
title: "gouwens mouse patch-seq analysis"
author: "Selen Bayram"
date: "2023-07-13"
output: rmarkdown::html_vignette
---

```{r setup, include = FALSE}
knitr::opts_knit$set(root.dir = "/nethome/kcni/sbayram/gouwens_test/")
```

## Setup
# Load packages
```{r}
#install.packages("tidyverse")
library(tidyverse)
#install.packages("Seurat")
library(Seurat)
#install.packages("janitor")
library(janitor)
library(stringr)
```

## Load metadata and count matrix
```{r}
mouse_count_matrix <- read.csv("/external/rprshnas01/netdata_kcni/stlab/Public/AIBS_patchseq_2020/mouse/transcriptomics/20200513_Mouse_PatchSeq_Release_count.csv", row.names = 1)

# column 19 include the transcriptomics labels so we set the row names in metadata to match the count matrix
mouse_meta_data <- read.csv("/external/rprshnas01/netdata_kcni/stlab/Public/AIBS_patchseq_2020/mouse/20200711_patchseq_metadata_mouse.csv", row.names = 19)
```

## Clean up metadata
```{r}
# replace the white space in the column names with underscore
# case = "snake" will convert the column names to snake case (e.g. "T-type" to "t_type")
mouse_meta_data <- mouse_meta_data %>% clean_names(case = "snake")
mouse_count_matrix <- mouse_count_matrix %>% clean_names(case = "snake")

# convert the row names to snake case
# unfortunately, there was no function that i could find so i had to do the following:
mouse_meta_data <- mouse_meta_data %>% 
    rownames_to_column("cell_id") %>% 
    mutate(cell_id = str_to_lower(cell_id)) %>% 
    mutate(cell_id = str_replace_all(cell_id, "-", "_")) %>%
    column_to_rownames("cell_id")
```


## Add a column to the metadata to indicate the cell subclass_label
```{r}
# create a new column called "subclass_label" in the metadata
## mouse_meta_data <- mouse_meta_data %>% mutate(subclass_label = word(mouse_meta_data$x, 1))
# change the empty cells to NA
## mouse_meta_data["subclass_label"][mouse_meta_data["subclass_label"] == ""] <- "NA"

# rename the column "x" to "subclass_label"
mouse_meta_data <- mouse_meta_data %>%
    rename("subclass_label" = "x")

# extract the first word from the subclass_label column
mouse_meta_data <- mouse_meta_data %>%
    mutate(subclass_label = word(subclass_label, 1))

# change the empty cells to NA
mouse_meta_data["subclass_label"][mouse_meta_data["subclass_label"] == ""] <- "NA"

# remove the rows with NA in subclass_label
mouse_meta_data <- mouse_meta_data %>%
    filter(subclass_label != "NA")
```

## Create Seurat object
```{r}
seu_mouse <- CreateSeuratObject(counts = mouse_count_matrix, meta.data = mouse_meta_data)
Idents(seu_mouse)
```

## Processing
# Normalization
```{r}
# scale.factor = 1000000 so that we get counts per million (CPM)
seu_mouse <- NormalizeData(seu_mouse, normalization.method = "LogNormalize", scale.factor = 1000000)
```

# Find variable features-genes-
```{r}
seu_mouse <- FindVariableFeatures(seu_mouse, selection.method = "vst", nfeatures = 2000)
```

# Scale data
```{r}
seu_mouse <- ScaleData(seu_mouse)
```

# Run PCA
```{r}
seu_mouse <- RunPCA(seu_mouse, npcs = 50)
```

# Use ElbowPlot to determine the number of PCs to use in clustering
```{r}
ElbowPlot(seu_mouse, ndims = 50)
```

# Run PCA again with the number of PCs determined by ElbowPlot
```{r}
# we could use 13-20 PCs
seu_mouse <- RunPCA(seu_mouse, npcs = 20)
```

# Find neighbors
```{r}  
seu_mouse <- FindNeighbors(seu_mouse, reduction = "pca", dims = 1:20)
```

# Find clusters
```{r}
seu_mouse <- FindClusters(seu_mouse, resolution = 0.5)
```

# Show the number of cells in each cluster
```{r}
table(seu_mouse$seurat_clusters)
table(seu_mouse$seurat_clusters, seu_mouse$subclass_label)
```

# Run UMAP
```{r}
seu_mouse <- RunUMAP(seu_mouse, reduction = "pca", dims = 1:20)
```

## Visualizations
# DimPlot
```{r}
seurat_umap <- DimPlot(seu_mouse, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE)
# seurat_pca <- DimPlot(seu_mouse, reduction = "pca", group.by = "seurat_clusters", label = TRUE, repel = TRUE)
subclass_umap <- DimPlot(seu_mouse, reduction = "umap", group.by = "subclass_label", label = TRUE, repel = TRUE)
seurat_umap + subclass_umap
```
