---
title: "Panoramics Tutorial"
author: "Selen Bayram"
date: "2023-06-14"
output: html_document
---

#Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(Seurat)
```

```{r}
hum_meta <- read.csv("AIBS_human_meta_mini.csv", row.names = 1)
hum_counts <- read.csv("AIBS_human_counts_mini.csv", row.names = 1)

hum_counts_sparse <- as(as.matrix(hum_counts), "sparseMatrix")


row.names(hum_meta) <- hum_meta$sample_name

hum_meta
```

# Create Seurat object
```{r}

hum_counts_sparse_transposed = as(as.matrix(t(hum_counts)), "sparseMatrix")
Seu_hum <- CreateSeuratObject(counts = hum_counts_sparse_transposed, meta.data = hum_meta, assay = "RNA")

#rm(hum_counts, hum_meta)
```

active identity:
```{r}
Idents(Seu_hum)
Seu_hum@active.ident
```

# Normalization of data
Counts per million:
```{r}
Seu_hum <- NormalizeData(Seu_hum, normalization.method = "LogNormalize", 
                         scale.factor = 1000000)
Seu_hum@assays$RNA@data #access the normalized data
```

Find variable features (genes)
```{r}
Seu_hum <- FindVariableFeatures(Seu_hum, selection.method = "vst", 
                                nfeatures = 2000)

Seu_hum@assays$RNA@var.features
```

Scale data:
```{r}
Seu_hum <- ScaleData(Seu_hum)
```

Run PCA:
```{r}
Seu_hum <- RunPCA(Seu_hum, npcs = 50)
ElbowPlot(Seu_hum, ndims = 50)
```

Find Neighbors:
```{r}
Seu_hum <- FindNeighbors(Seu_hum, reduction = "pca", dims = 1:20)
```

Find Clusters:
```{r}
Seu_hum <- FindClusters(Seu_hum, resolution = 0.5)
```

```{r}
table(Seu_hum$seurat_clusters)
table(Seu_hum$seurat_clusters, Seu_hum$subclass_label)
```

Run UMAP:
```{r}
Seu_hum <- RunUMAP(Seu_hum, reduction = "pca", dims = 1:20)

DimPlot(Seu_hum, reduction = "umap", group.by = "seurat_clusters", label = TRUE,
        repel = TRUE)
```

```{r}
DimPlot(Seu_hum, reduction = "umap", group.by = "subclass_label", label = TRUE)
```

Get marker genes for all clusters:
```{r}
all_clusters_markers = FindAllMarkers(Seu_hum, logfc.threshold = log(2), 
                                      min.pct = 0.5)
```


```{r}
features = cluster_3_vs_all %>%
  arrange(p_val_adj) %>%
  head(n = 6) %>%
  row.names()
```

# Violin Plot
```{r}
#svg("cluster_3_violin_plot.svg")
#png("cluster_3_violin_plot.svg", res = 300)
VlnPlot(Seu_hum, features = features)
#dev.off()
```

# Feature Plot
```{r}
FeaturePlot(Seu_hum, features = features)
```

# Dot Plot
```{r}
DotPlot(Seu_hum, features = features) + RotatedAxis()
```

# Heatmap
```{r}
# subsetting 100 cells per cluster from the data of 2500 cells
DoHeatmap(subset(Seu_hum, downsample = 100), features = features, size = 3, 
          slot = "data") + scale_fill_viridis_c()
# scale_fill_viridis_c() changes the colour palette from ugly pink to black
```

# Is there any difference between clusters based on the sex of the cell donors?
```{r}
Seu_hum@meta.data <- Seu_hum@meta.data %>%
  mutate(sex_cluster3 = paste(donor_sex_label, seurat_clusters, sep = "_"))

table(Seu_hum$sex_cluster3)
```

```{r}
Idents(Seu_hum) <- "sex_cluster3"

sex_cluster3_markers <- FindMarkers(Seu_hum, ident.1 = "F_3", ident.2 = "M_3",
                                    logfc.threshold = log(2), min.pct = 0.5)
```

# genes that are expressed differently between male vs female
```{r}
sex_cluster3_markers %>%
  arrange(p_val_adj)
# genes that are upregulated in females compared to males are +ve values in the 
# table since ident.1 == F (if ident.1 and ident.2 switched, values would be -ve)
```
