---
title: 'KCNI Summer Academy 2023: Day 3'
author: "Selen Bayram"
date: "11/07/2023"
output: pdf_document
---

# Dataset integration & automated cell type annotation
## Setup
### Set working directory
```{r setup}
knitr::opts_knit$set(root.dir = "/Users/selenbayram/triplab/KCNI_Summer_2023/")
```

### Load packages
```{r}
#install.packages("BiocManager")
#BiocManager::install("multtest")
library(multtest)
#BiocManager::install("limma")
library(limma)
#install.packages("metap")
library(metap)  

library(tidyverse)
library(Seurat)
```

```{r}
# install.packages("httpgd")
library(httpgd)
```

```{r}
# load smart
meta_smart <- read.csv("/Users/selenbayram/triplab/KCNI_Summer_2023/AIBS_smart_meta_mini.csv", row.names = 1) %>% mutate(method = "smart")
counts_smart <- read.csv("/Users/selenbayram/triplab/KCNI_Summer_2023/AIBS_smart_counts_mini.csv", row.names = 1)
```

```{r}
# load 10x
meta_10x <- read.csv("/Users/selenbayram/triplab/KCNI_Summer_2023/SEA-AD_10x_meta_mini.csv", row.names = 1) %>% mutate(method = "10x")
counts_10x <- read.csv("/Users/selenbayram/triplab/KCNI_Summer_2023/SEA-AD_10x_counts_mini.csv", row.names = 1)
```

Filtering the count matrices
```{r}
# finding the list of genes shared between the two datasets
common_genes <- intersect(colnames(counts_smart), colnames(counts_10x))
```

```{r}
# use shared genes to subset count matrices
# indexing by columns [,common_genes] and by rows [common_genes,]
# only columns that are kept are the ones that are in common_genes

counts_smart <- counts_smart[,common_genes]
counts_10x <- counts_10x[,common_genes]
```

Creating Seurat objects
```{r}
seu_smart_intg <- CreateSeuratObject(counts = t(counts_smart), meta.data = meta_smart)
seu_10x_intg <- CreateSeuratObject(counts = t(counts_10x), meta.data = meta_10x)
dim(counts_10x)
dim(counts_smart)
```

```{r}
# remove the large dataframes from the environment
rm(counts_smart, counts_10x, meta_10x, meta_smart)
```

Preprocessing for Integration 
```{r}
seu_list <- c(seu_smart_intg, seu_10x_intg)
# rm(seu_smart_intg, seu_10x_intg)
```

```{r}
# apply the function() to each seurat object in seu_list
# SCTransform() is a function from Seurat that performs normalization and scaling
seu_list <- lapply(X = seu_list, FUN = function(x) {
  x <- SCTransform(x, variable.features.n = 3000, conserve.memory = TRUE)
})
```

```{r}
# select features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = seu_list, nfeatures = 3000)
seu_list <- PrepSCTIntegration(object.list = seu_list, anchor.features = features)
```

```{r}
# controlling for the differences between within-dataset variabilities
anchors <- FindIntegrationAnchors(object.list = seu_list, normalization.method = "SCT", anchor.features = features)
```

```{r}
# integrate the datasets
seu_intg <- IntegrateData(anchorset = anchors, normalization.method = "SCT")
```

```{r}
rm(seu_list, anchors, features, common_genes)
```

Initial Analysis
```{r}
DefaultAssay(seu_intg) <- "integrated"
```

```{r}
# scale the data
seu_intg <- ScaleData(seu_intg, verbose = FALSE)

# run PCA
seu_intg <- RunPCA(seu_intg, npcs = 40, verbose = FALSE)
```

```{r}
# elbow plot
ElbowPlot(seu_intg, ndims = 40)
```

```{r}
seu_intg <- RunPCA(seu_intg, npcs = 30, verbose = FALSE)
```

```{r}
# find neighbors and clusters and run UMAP
seu_intg <- FindNeighbors(seu_intg, reduction = "pca", dims = 1:30) %>% 
    FindClusters(resolution = 0.5) %>% # current active identity is set to "seurat_clusters" here
    RunUMAP(reduction = "pca", dims = 1:30)
```

```{r}
# numbers 0-24 represent the clusters and each number underneath is the number of cells grouped into that cluster
table(seu_intg$seurat_clusters)
```

Visualization
```{r}
Idents(seu_intg) <- "seurat_clusters"
```

```{r}
DimPlot(seu_intg, reduction = "umap", label = TRUE)
```
```{r}
DimPlot(seu_intg, reduction = "umap", group.by = "method", shuffle = TRUE)
```

```{r}
DimPlot(seu_intg, reduction = "umap", group.by = "subclass_label", 
        split.by = "method", label=TRUE)
```
    
```{r}
# change the labels to uppercase so that the naming convention between the two datasets is consistent
seu_intg@meta.data$subclass_label <- toupper(seu_intg@meta.data$subclass_label)
```

```{r}
DimPlot(seu_intg, reduction = "umap", group.by = "subclass_label", 
        split.by = "method", label=TRUE)
```

```{r}
DimPlot(seu_intg, reduction = "umap", group.by = "subclass_label", 
        cells.highlight = list(VIP = seu_intg@meta.data %>% 
                                filter(subclass_label == "VIP") %>% 
                                base::row.names(.),
                               SST = seu_intg@meta.data %>% 
                                filter(subclass_label == "SST") %>% 
                                base::row.names(.)),
        cols.highlight = c("slateblue", "cadetblue"))
```

```{r}
# save the integrated dataset as seurat object
# saveRDS(seu_intg, file = "seu_intg.rds")
```

```{r}
FeaturePlot(seu_intg, features = c("VIP", "PVALB", "SST"), split.by = "method")
```

Find Conserved Markers
```{r}
DefaultAssay(seu_intg) <- "SCT"
```

= optional: parallel processing =
```{r}
install.packages("future")
library(future)

availableCores()
```

```{r}
plan("multisession", workers = 6) # changes from sequential to parallel processing
```

```{r}
Idents(seu_intg) <- "seurat_clusters"
```

```{r}
# chandelier cells are consistently marked by these genes across the two datasets
cluster_21_markers <- FindConservedMarkers(seu_intg, ident.1 = 21, grouping.var = "method", logfc.threshold = log(2), minpct = 0.25)
```

```{r}
# cluster_21_markers 
# dim(cluster_21_markers)
```

```{r}
# back to sequential processing bc we are done with FindConservedMarkers()
plan("sequential")
```

```{r}
DefaultAssay(seu_intg) <- "integrated"
```

```{r}
seu_intg$cluster_method <- paste(seu_intg$seurat_clusters, seu_intg$method, sep = "_")
```

```{r}
table(seu_intg$cluster_method)
```

```{r}
Idents(seu_intg) <- "cluster_method"
```

```{r}
# looking at differential expression between the two datasets // cluster 21: chandelier cells
cluster_21_method_de <- FindMarkers(seu_intg, ident.1 = "21_10x", ident.2 = "21_smart", logfc.threshold = log(2), min.pct = 0.25)
```

```{r}
# cluster_21_method_de
```
```{r}
# looking at differential expression between the two datasets // cluster 10: PVALB cells
cluster_10_method_de <- FindMarkers(seu_intg, ident.1 = "10_10x", ident.2 = "10_smart", logfc.threshold = log(2), min.pct = 0.25)
```
```{r}
# cluster_10_method_de
```

```{r}
cluster_10_top_de <- cluster_10_method_de %>% 
    # arrange(avg_log2FC) %>%
    head(3) %>% 
    row.names(.) 

cluster_10_method_de %>% 
    arrange(desc(avg_log2FC)) %>%
    head(3) %>% 
    row.names(.) 
```
```{r}
FeaturePlot(seu_intg, features = c("PCDH", "CA8", "FRMPD4", "ITGAX", "THSD4", "ICAM1"), 
            split.by = "method")
```

```{r}
Idents(seu_intg) <- "seurat_clusters"
```

Comparing DotPlot() when using different assays
```{r}
DefaultAssay(seu_intg) <- "SCT"
p_sct <- DotPlot(seu_intg, features = cluster_10_top_de, idents = 10, group.by = "method") + RotatedAxis()
DefaultAssay(seu_intg) <- "RNA"
p_rna <- DotPlot(seu_intg, features = cluster_10_top_de, idents = 10, group.by = "method") + RotatedAxis()
DefaultAssay(seu_intg) <- "integrated"
p_intg <- DotPlot(seu_intg, features = cluster_10_top_de, idents = 10, group.by = "method") + RotatedAxis()
```

```{r}
p_sct + p_rna + p_intg
```

Automating Cell Type Identification
TransferData
Project the PCA structure of reference dataset on to query dataset
```{r}
seu_smart <- readRDS("/Users/selenbayram/triplab/KCNI_Summer_2023/seu_smart_2.rds")
seu_10x <- readRDS("/Users/selenbayram/triplab/KCNI_Summer_2023/seu_10x_2.rds")
```

```{r}
seu_10x@meta.data$subclass_label <- toupper(seu_10x@meta.data$subclass_label)
seu_smart@meta.data$subclass_label <- toupper(seu_smart@meta.data$subclass_label)
```
```{r}
anchors <- FindTransferAnchors(reference = seu_10x, query = seu_smart, reference.reduction = "pca", dims = 1:30)
```

```{r}
predictions <- TransferData(anchorset = anchors, refdata = seu_10x$subclass_label, dims = 1:30)
```

```{r}
smart_query <- seu_smart %>% 
    AddMetaData(metadata = predictions)
```

```{r}
# add a new column with values T or F depending on whether the predicted label matches the true label
smart_query$prediction_match <- smart_query$subclass_label == smart_query$predicted.id
```

```{r}
table(smart_query$prediction_match)
table(smart_query$predicted.id)

# for the cekl types that only exit in reference dataset (10x) we expect all False by default
table(smart_query$predicted.id, smart_query$prediction_match)
```


MapQuery
projecting the query dataset onto the UMAP structure of the reference
```{r}
anchors <- FindTransferAnchors(reference = seu_10x, query = seu_smart, reference.reduction = "pca", dims = 1:30)
```

```{r}
seu_10x <- RunUMAP(seu_10x, return.model = TRUE, dims = 1:30)
seu_smart <- RunUMAP(seu_smart, return.model = TRUE, dims = 1:30)
```

```{r}
smart_query_2 <- MapQuery(anchorset = anchors, reference = seu_10x, query = seu_smart, refdata = list(subclass_label = "subclass_label"), reference.reduction = "pca", reduction.model = "umap")
```

```{r}
p3 <- DimPlot(smart_query_2, reduction = "ref.umap", group.by = "predicted.subclass_label", 
                label = TRUE, label.size = 3, repel = TRUE) + NoLegend() + 
                ggtitle("Query transferred labels")
p4 <- DimPlot(seu_10x, reduction = "umap", group.by = "subclass_label", 
                label = TRUE, label.size = 3, repel = TRUE) + NoLegend() + 
                ggtitle("Reference annotations")
```

```{r}
p3 + p4
```

```{r}
smart_query_2$predicted.subclass_label.score
VlnPlot(smart_query_2, group.by = "predicted.subclass_label", features = "predicted.subclass_label.score") + NoLegend()
```

```{r}
# Generate the numbers from 1 to 10
x <- 1:10

# Create a plot with x as the x-axis and x as the y-axis
plot(x, x, type = "l", lwd = 2, xlab = "Number", ylab = "Number", main = "Plot of Numbers 1-10")

```